Quotients of the magmatic operad:\\ lattice structures and convergent rewrite systems

### 2018-09-11 CAs_operads.tex
MAJ

Let, for any integer $\gamma \geq 1$, the binary trees $\LComb{\gamma}$
and $\RComb{\gamma}$ be respectively the left and the right combs of
degree $\gamma$. These trees are depicted as
\begin{equation}
    \LComb{\gamma} = \enspace
    \begin{tikzpicture}[xscale=.26,yscale=.3,Centering]
        \node(0)at(0.00,-5.25){};
        \node(2)at(2.00,-5.25){};
        \node(4)at(4.00,-3.50){};
        \node(6)at(6.00,-1.75){};
        \node[NodeST](1)at(1.00,-3.50){\begin{math}\Product\end{math}};
        \node[NodeST](3)at(3.00,-1.75){\begin{math}\Product\end{math}};
        \node[NodeST](5)at(5.00,0.00){\begin{math}\Product\end{math}};
        \draw[Edge](0)--(1);
        \draw[Edge,dotted](1)edge[]node[font=\tiny]{
            \begin{math}\gamma\! -\! 1\end{math}\hspace*{.6cm}}(3);
        \draw[Edge](2)--(1);
        \draw[Edge](3)--(5);
        \draw[Edge](4)--(3);
        \draw[Edge](6)--(5);
        \node(r)at(5.00,1.5){};
        \draw[Edge](r)--(5);
    \end{tikzpicture}
    \qquad \mbox{and} \qquad
    \RComb{\gamma} =
    \begin{tikzpicture}[xscale=.26,yscale=.3,Centering]
        \node(0)at(0.00,-1.75){};
        \node(2)at(2.00,-3.50){};
        \node(4)at(4.00,-5.25){};
        \node(6)at(6.00,-5.25){};
        \node[NodeST](1)at(1.00,0.00){\begin{math}\Product\end{math}};
        \node[NodeST](3)at(3.00,-1.75){\begin{math}\Product\end{math}};
        \node[NodeST](5)at(5.00,-3.50){\begin{math}\Product\end{math}};
        \draw[Edge](0)--(1);
        \draw[Edge](2)--(3);
        \draw[Edge](3)--(1);
        \draw[Edge](4)--(5);
        \draw[Edge,dotted](5)edge[]node[font=\tiny]{
            \hspace*{.6cm}\begin{math}\gamma\! -\! 1\end{math}}(3);
        \draw[Edge](6)--(5);
        \node(r)at(1.00,1.5){};
        \draw[Edge](r)--(1);
    \end{tikzpicture}\,,
\end{equation}
where the values on the dotted edges denote the number of internal
nodes they contain. We shall employ this drawing convention in the
sequel, combined with the convention stipulating that dotted edges
with no value have any number of internal nodes. Let us now define for
any $\gamma \geq 1$ the \Def{$\gamma$-comb associative operad}
$\CAs{\gamma}$ as the quotient operad $\Mag/_{\CongrCAs{\gamma}}$ where
$\CongrCAs{\gamma}$ is the smallest operad congruence of $\Mag$
satisfying
\begin{equation} \label{equ:congruence_CAs_gamma}
    \LComb{\gamma} \enspace \CongrCAs{\gamma} \enspace \RComb{\gamma}.
\end{equation}
Notice that $\CongrCAs{1}$ is trivial so that
$\CAs{1} = \Mag$, and that $\CongrCAs{2}$ is the operad congruence
defining $\As$ so that $\CAs{2} = \As$. Let also
\begin{equation}
    \CAsAll := \left\{\CAs{\gamma} : \gamma\geq 1\right\}
\end{equation}
be the set of all the $\gamma$-comb associative operads.
\medbreak

>>>>

As in Section~\ref{sec:Magmatic_operads}, the left and the right combs of
degree $\gamma$ are denoted by $\LComb{\gamma}$ and $\RComb{\gamma}$,
respectively. In the sequel , we shall employ the drawing convention
introduced after~\eqref{equ:left_and_right_combs}: the values on dotted
edges in a binary tree denote the number of internal nodes they contain.
Moreover, we also employ the convention stipulating that dotted edges
with no value have any number of internal nodes. Let us now define for
any $\gamma \geq 1$ the \Def{$\gamma$-comb associative operad}
$\CAs{\gamma}$ as the quotient operad $\Mag/_{\CongrCAs{\gamma}}$ where
$\CongrCAs{\gamma}$ is the smallest operad congruence of $\Mag$
satisfying
\begin{equation} \label{equ:congruence_CAs_gamma}
    \LComb{\gamma} \enspace \CongrCAs{\gamma} \enspace \RComb{\gamma}.
\end{equation}
Notice that $\CongrCAs{1}$ is trivial so that
$\CAs{1} = \Mag$, and that $\CongrCAs{2}$ is the operad congruence
defining $\As$ so that $\CAs{2} = \As$. Let also
\begin{equation}
    \CAsAll := \left\{\CAs{\gamma} : \gamma\geq 1\right\}
\end{equation}
be the set of all the $\gamma$-comb associative operads.
\medbreak

### 2018-09-11 Mag_quotients.tex
MAJ

We terminate this section with an example illustrating the lattice
constructions on $\QMag$ in the case where the
characteristic of the field $\K$ is different from $2$. For that, we
introduce various operads, which requires the following notations, also
used in Section~\ref{sec:CAs_d}. Let, for any integer $\gamma \geq 1$,
the binary trees $\LComb{\gamma}$ and $\RComb{\gamma}$ be respectively
the left and the right combs of degree $\gamma$. These trees are depicted
as
\begin{equation}
    \LComb{\gamma} = \enspace
    \begin{tikzpicture}[xscale=.26,yscale=.3,Centering]
        \node(0)at(0.00,-5.25){};
        \node(2)at(2.00,-5.25){};
        \node(4)at(4.00,-3.50){};
        \node(6)at(6.00,-1.75){};
        \node[NodeST](1)at(1.00,-3.50){\begin{math}\Product\end{math}};
        \node[NodeST](3)at(3.00,-1.75){\begin{math}\Product\end{math}};
        \node[NodeST](5)at(5.00,0.00){\begin{math}\Product\end{math}};
        \draw[Edge](0)--(1);
        \draw[Edge,dotted](1)edge[]node[font=\tiny]{
            \begin{math}\gamma\! -\! 1\end{math}\hspace*{.6cm}}(3);
        \draw[Edge](2)--(1);
        \draw[Edge](3)--(5);
        \draw[Edge](4)--(3);
        \draw[Edge](6)--(5);
        \node(r)at(5.00,1.5){};
        \draw[Edge](r)--(5);
    \end{tikzpicture}
    \qquad \mbox{and} \qquad
    \RComb{\gamma} =
    \begin{tikzpicture}[xscale=.26,yscale=.3,Centering]
        \node(0)at(0.00,-1.75){};
        \node(2)at(2.00,-3.50){};
        \node(4)at(4.00,-5.25){};
        \node(6)at(6.00,-5.25){};
        \node[NodeST](1)at(1.00,0.00){\begin{math}\Product\end{math}};
        \node[NodeST](3)at(3.00,-1.75){\begin{math}\Product\end{math}};
        \node[NodeST](5)at(5.00,-3.50){\begin{math}\Product\end{math}};
        \draw[Edge](0)--(1);
        \draw[Edge](2)--(3);
        \draw[Edge](3)--(1);
        \draw[Edge](4)--(5);
        \draw[Edge,dotted](5)edge[]node[font=\tiny]{
            \hspace*{.6cm}\begin{math}\gamma\! -\! 1\end{math}}(3);
        \draw[Edge](6)--(5);
        \node(r)at(1.00,1.5){};
        \draw[Edge](r)--(1);
    \end{tikzpicture}\,,
\end{equation}
where the values on the dotted edges denote the number of internal
nodes they contain.

We define the \Def{anti-associative operad} by
\begin{equation}
  \label{equ:anti-associative_operad}
  \AAs:=\KMag/_{\IAAs}
\end{equation}
where $\IAAs$ is the ideal spanned by
\begin{math}
  \LComb{2}+\RComb{2}
\end{math}.
\Todo{Illustrer avec la représentation graphique ?}
Using the Buchberger algorithm for operads~\cite[Section 3.7]{DK10}, we
check that the set of rewrite rules
\begin{equation}
  \label{equ:convergent_presentation_of_AAS}
  \left\{\LComb{2}\Rew -\RComb{2},\
  \RComb{3}\Rew 0
  \right\}
\end{equation}
is a convergent presentation of $\AAs$. We point out that this statement
is false if the characteristic of $\K$ is equal to $2$. Moreover, using
the convergent presentation~\eqref{equ:convergent_presentation_of_AAS},
we have
  \begin{equation}
    \label{equ:Hilbert_series_of_AAS}
    \HilbertSeries_{\AAs}(t)=t+t^2+t³.
    \end{equation}
\medbreak

We define the \Def{2-nihilpotente operad} by
\begin{equation}
  \label{equ:2-nihilpotente_operad}
  \NiHil:=\KMag/_{\INiHil}
  \end{equation}
where $\INiHil$ is the ideal spanned by the two trees $\LComb{3}$ and
$\RComb{3}$.
\Todo{Est-ce qu'il faut rajouter le 2 dans la macro désignant cette
  opérade ? De plus, Samuele,\\
  pourrais-tu rajouter la référence de Loday ?}
We have
\begin{equation}
  \label{equ:Hilbert_series_of_NiHil}
  \HilbertSeries_{\NiHil}(t)=t+t^2.
\end{equation}
\Todo{Est-ce bien la bonne série ? Samuele, tu avais l'air de dire que
  c'était plutôt $t+2t^2$. Ou peut-être que NiHil est la $1$-
  nihilipotente opérade ?}

@@

\begin{Lemma}
  \label{lem:presentation_of_RC}
  We have an isomorphism
  \begin{equation}
  \label{equ:presentation_of_RC}
  \RC{\gamma}\simeq\KMag/_{\IRCC{\gamma}}
  \end{equation}
  where $\IRCC{\gamma}$ is the ideal spanned by the elements
  $\Tfr-\RComb{\gamma}$, where $\Tfr$ runs over all the binary trees of
  arity $\gamma+1$. In other words, $\RC{\gamma}$ is a combinatorial
  realization of
  \begin{math}
    \KMag/_{\IRCC{\gamma}}.
    \end{math}
\end{Lemma}

@@

\begin{proof}
  First, we show~\eqref{equ:example_lower-bound}. By definition of the
  lower-bound $\InfQMag$, the ideal of relations of
  $\KAs\InfQMag\AAs$ is spanned by the two elements
  $\LComb{3}-\RComb{3}$ and $\LComb{3}+\RComb{3}$. By linear
  transformations applied to these generators, the ideal of relations of
  $\KAs\InfQMag\AAs$ is spanned by $\LComb{3}$ and $\RComb{3}$,
  that is it is equal to $\INiHil$, which
  proves~\eqref{equ:example_lower-bound}.

  \smallbreak
  Now, we show~\eqref{equ:example_upper-bound}. We denote by
  \begin{math}
    \pi:\KMag\to\KAs\SupQMag\AAs
  \end{math}
  the natural projection. Let $\Tfr$ be a tree of arity $4$. From
  Lemma~\ref{lem:presentation_of_RC}, the kernel of $\pi$ is spanned by
  the elements $\alpha_{\Tfr}:=\Tfr-\RComb{3}$, where $\Tfr$ is a binary
  tree of arity $4$. The elements $\alpha_{\Tfr}$ belong to the ideal
  $\IAs$ of relations of $\KAs$ and to the ideal $\IAAs$ since both
  $[\Tfr]_{\IAAs}$ and $[\RComb{3}]_{\IAAs}$ are equal to $[0]_{\IAAs}$.
  The last statement is shown using the convergent
  presentation~\eqref{equ:convergent_presentation_of_AAS}. Hence, the kernel
  of $\pi$ is included in $\IAs\cap\IAAs$, so that, by definition of the
  upper-bound $\SupQMag$, $\pi$ induces a surjective operad morphism
  \begin{math}
    \overline{\pi}:\As\SupQMag\AAs\to\RC{3}
  \end{math}.
  We conclude using a dimension argument: by definition, the Hilbert series
  of $\RC{3}$ is equal to
  \begin{equation}
    \label{equ:Hilbert_series_of_RC3}
    t+t^2+2t^3+\sum_{n\geq 4}t^n
  \end{equation}
  and from the Grassmann formula analog given in
  Theorem~\ref{thm:Grassmann_formula_for_Hilbert_series_of_QMag}, we check
  that the Hilbert series of $\As\SupQMag\AAs$ is also equal
  to~\eqref{equ:Hilbert_series_of_RC3}. Hence, $\overline{\pi}$ is an
  isomorphism, which proves~\eqref{equ:example_upper-bound}.
\end{proof}
\medbreak

>>>>

We terminate this section with an example illustrating the lattice
constructions on $\QMag$ in the case where the
characteristic of the field $\K$ is different from $2$. For that, we
introduce various operads, which requires the following notations, also
used in Section~\ref{sec:CAs_d}. Let, for any integer $\gamma \geq 1$,
the binary trees $\LComb{\gamma}$ and $\RComb{\gamma}$ be respectively
the left and the right combs of degree $\gamma$. These trees are depicted
as
\begin{equation}
  \label{equ:left_and_right_combs}
    \LComb{\gamma} = \enspace
    \begin{tikzpicture}[xscale=.26,yscale=.3,Centering]
        \node(0)at(0.00,-5.25){};
        \node(2)at(2.00,-5.25){};
        \node(4)at(4.00,-3.50){};
        \node(6)at(6.00,-1.75){};
        \node[NodeST](1)at(1.00,-3.50){\begin{math}\Product\end{math}};
        \node[NodeST](3)at(3.00,-1.75){\begin{math}\Product\end{math}};
        \node[NodeST](5)at(5.00,0.00){\begin{math}\Product\end{math}};
        \draw[Edge](0)--(1);
        \draw[Edge,dotted](1)edge[]node[font=\tiny]{
            \begin{math}\gamma\! -\! 1\end{math}\hspace*{.6cm}}(3);
        \draw[Edge](2)--(1);
        \draw[Edge](3)--(5);
        \draw[Edge](4)--(3);
        \draw[Edge](6)--(5);
        \node(r)at(5.00,1.5){};
        \draw[Edge](r)--(5);
    \end{tikzpicture}
    \qquad \mbox{and} \qquad
    \RComb{\gamma} =
    \begin{tikzpicture}[xscale=.26,yscale=.3,Centering]
        \node(0)at(0.00,-1.75){};
        \node(2)at(2.00,-3.50){};
        \node(4)at(4.00,-5.25){};
        \node(6)at(6.00,-5.25){};
        \node[NodeST](1)at(1.00,0.00){\begin{math}\Product\end{math}};
        \node[NodeST](3)at(3.00,-1.75){\begin{math}\Product\end{math}};
        \node[NodeST](5)at(5.00,-3.50){\begin{math}\Product\end{math}};
        \draw[Edge](0)--(1);
        \draw[Edge](2)--(3);
        \draw[Edge](3)--(1);
        \draw[Edge](4)--(5);
        \draw[Edge,dotted](5)edge[]node[font=\tiny]{
            \hspace*{.6cm}\begin{math}\gamma\! -\! 1\end{math}}(3);
        \draw[Edge](6)--(5);
        \node(r)at(1.00,1.5){};
        \draw[Edge](r)--(1);
    \end{tikzpicture}\,,
\end{equation}
where the values on the dotted edges denote the number of internal
nodes they contain.

We first recall that the \Def{linear associative operad} is
\begin{equation}
  \label{equ:linear_associative_operad}
  \KAs:=\KMag/_{\IAs}
\end{equation}
where $\IAs$ is the ideal spanned
\begin{math}
  \LComb{2}-\RComb{2},
\end{math}
and that its Hilbert series is
\Todo{référence ?}
\begin{equation}
    \label{equ:Hilbert_series_of_As}
    \HilbertSeries_{\KAs}(t)=\sum_{n\geq 1}t^n,
    \end{equation}

We define the \Def{anti-associative operad} by
\begin{equation}
  \label{equ:anti-associative_operad}
  \AAs:=\KMag/_{\IAAs}
\end{equation}
where $\IAAs$ is the ideal spanned by
\begin{math}
  \LComb{2}+\RComb{2}
\end{math}.
Using the Buchberger algorithm for operads~\cite[Section 3.7]{DK10}, we
check that the set of rewrite rules
\begin{equation}
  \label{equ:convergent_presentation_of_AAS}
  \left\{\LComb{2}\Rew -\RComb{2},\
  \RComb{3}\Rew 0
  \right\}
\end{equation}
is a convergent presentation of $\AAs$. We point out that this statement
is false if the characteristic of $\K$ is equal to $2$. Moreover, using
the convergent presentation~\eqref{equ:convergent_presentation_of_AAS},
we have
  \begin{equation}
    \label{equ:Hilbert_series_of_AAs}
    \HilbertSeries_{\AAs}(t)=t+t^2+t³.
    \end{equation}
\medbreak

We define the \Def{2-nihilpotente operad} by
\begin{equation}
  \label{equ:2-nihilpotente_operad}
  \NiHil:=\KMag/_{\INiHil}
  \end{equation}
where $\INiHil$ is the ideal spanned by the two trees $\LComb{2}$ and
$\RComb{2}$.
\Todo{Est-ce qu'il faut rajouter le 2 dans la macro désignant cette
  opérade ? De plus, Samuele,\\
  pourrais-tu rajouter la référence de Loday ?}
We have
\begin{equation}
  \label{equ:Hilbert_series_of_NiHil}
  \HilbertSeries_{\NiHil}(t)=t+t^2.
\end{equation}
\Todo{Est-ce bien la bonne série ? Samuele, tu avais l'air de dire que
  c'était plutôt $t+2t^2$. Ou peut-être que NiHil est la $1$-
  nihilipotente opérade ?}

@@

\begin{Lemma}
  \label{lem:presentation_of_RC}
  We have an isomorphism
  \begin{equation}
  \label{equ:presentation_of_RC}
  \RC{\gamma}\simeq\KMag/_{\IRC{\gamma}}
  \end{equation}
  where $\IRC{\gamma}$ is the ideal spanned by the elements
  $\Tfr-\RComb{\gamma}$, where $\Tfr$ runs over all the binary trees of
  arity $\gamma+1$. In other words, $\RC{\gamma}$ is a combinatorial
  realization of
  \begin{math}
    \KMag/_{\IRC{\gamma}}.
    \end{math}
\end{Lemma}

@@

\begin{proof}
  First, we show~\eqref{equ:example_lower-bound}. The ideal of relations
  of $\KAs\InfQMag\AAs$ is equal to $\IAs+\IAAs$, so that it is spanned
  by the two elements $\LComb{2}-\RComb{2}$ and $\LComb{2}+\RComb{2}$. By
  linear transformations applied to these generators, $\IAs+\IAAs$ is
  spanned by $\LComb{2}$ and $\RComb{2}$, that is it is equal to
  $\INiHil$, which proves~\eqref{equ:example_lower-bound}.

  \smallbreak
  Now, we show~\eqref{equ:example_upper-bound}. We denote by
  \begin{math}
    \pi:\KMag\to\KAs\SupQMag\AAs
  \end{math}
  the natural projection. Let $\Tfr$ be a tree of arity $4$ and let
  us define $\alpha_{\Tfr}:=\Tfr-\RComb{3}$. The elements $\alpha_{\Tfr}$
  belong to $\IAs$ and to $\IAAs$ since both $[\Tfr]_{\IAAs}$ and to
  $[\RComb{3}]_{\IAAs}$ are equal to $[0]_{\IAAs}$. The last statement is
  shown using the convergent
  presentation~\eqref{equ:convergent_presentation_of_AAS}. Hence, the
  ideal generated by the elements $\alpha_{\Tfr}$ is included in 
  $\IAs\cap\IAAs=\ker(\pi)$, so that $\pi$ induces a surjective morphism
  \begin{math}
    \overline{\pi}:\RC{3}\to\KAs\SupQMag\AAs
  \end{math}
  from Lemma~\ref{lem:presentation_of_RC}. We conclude using Hilbert
  series: $\HilbertSeries_{\KAs\SupQMag\AAs}(t)$ is computed
  using the Grassmann formula analog with
  Formulas~\eqref{equ:Hilbert_series_of_As},
  \eqref{equ:Hilbert_series_of_AAs} and
  ~\eqref{equ:Hilbert_series_of_NiHil} and it turns out to be equal to
  $\HilbertSeries_{\RC{3}}(t)$ which is given
  in~\eqref{equ:Hilbert_series_of_RC}. Hence, $\overline{\pi}$ is an
  isomorphism, which proves~\eqref{equ:example_upper-bound}.
\end{proof}
\medbreak
